An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-07-10 main.tex
Various edits

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a large variety of strategic interactions. Each player can choose
between cooperation (C) or defection (D). The decisions are made simultaneously
and independently. The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This extended form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as zero determinant (ZD) strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings (more details given in
Section~\ref{sec:the_moran_process}).
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, stewart2013extortion}, however these often make simplifying assumptions
and/or do not consider sophisticated behaviour: only considering strategies that
either cooperate or defect, or are limited to classes of strategies
such as memory-one strategies that only use the previous round of play.

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 40 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}).

@@

Using the Axelrod library and the many strategies it contains, we obtain
fixation probabilities for all pairs of strategies, identifying
those that are effective invaders and those resistant to invasion, for
population sizes $N=2$ to $N=14$. Moreover we present a number of strategies
that were created via reinforcement algorithms (evolutionary and particle
swarm algorithms) that are among the best invaders and resistors of invasion
known to date, and show that handshaking mechanims naturally arise from these
processes as an invasion-resistance mechanism.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider non-sophisticated strategies. Some work has
looked at evolutionary stability of strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

@@

To carry out this large numerical experiment, \input{./tex/num_strategies}
strategies, listed (with their properties) in Appendix~\ref{app:list_of_players},
are used from the Axelrod library. There are
\input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it differs from TF1 and TF2
in that it lacks a handshake mechanism.

TF2 always starts with CD and will defect against opponents that start with
DD\@. It plays CDD against itself and then cooperates thereafter. Cooperation
can be rescued after a failed handshake by a complex sequence of plays
which eventually results in mutual cooperation with Firm but Fair, Fortress3,
Fortress4, and Grofman (always) and Evolved HMM 5 and GTFT (depending on the
random seed). TF2 defects against all other players in the study, barring
unusual cases arising from particular randomizations. Figure~\ref{fig:tf2} shows
all 16 states of the strategy used for training although it is evident that
states 6 and 7 are not reachable. They are nonetheless presented as they are
outputs of the training algorithm.

@@

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. As it has little potential
        interaction with itself (recall that \(N=2\)), its
        aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while handshake will not.
\end{enumerate}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/average_rank_vs_population_size_invade.pdf}
    \caption{Ranks of all strategies according to \(x_1\) for different
    population sizes.}
    \label{fig:ranks_v_size_invade}
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/average_rank_vs_population_size_resist.pdf}
    \caption{Ranks of all strategies according to \(x_{N-1}\) for different
    population sizes.}
    \label{fig:ranks_v_size_resist}
\end{figure}

@@

As described in Section~\ref{sec:strong_invaders} the strategies trained with
the payoff maximizing objective are among the best invaders in the library
however they are not as resistant to invasion as the strategies trained using a
Moran objective function. These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms.

@@

For all starting populations
\(i\in\{1, N/2, N-1\}\) the ranks of strategies are relatively stable across the
different values of \(N>2\) however for \(N=2\) there is a distinct difference.
This highlights that there is little that can be inferred about the evolutionary
performance of a strategy in a large population from its performance in a small
population.

This is confirmed by the performance of the zero determinant strategies. While
some do rank relatively highly for \(N=2\) (ZD-extort-4 has rank 16) this rank
does not translate to larger populations.

@@

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than C necessarily involves
a defection. (A handshake consisting of a longer sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders still need to cooperate with themselves,
and so in the absense of a handshake mechanism or knowledge of the population
distribution, a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping dramatically
as the population size increases.

@@

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only revealed for CS (a human designed strategy) but also for
two FSM strategies (TF1 and TF2) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. Figure~\ref{fig:cooperation_rates} shows the cooperation
rate of TF1, TF2, TF3 and CS
for each round of a match against all the opponents in this study. Whilst TF3
does not have a strict handshake mechanism it is clear that all these strategies
start a match by cooperating. It is then evident that TF3 cooperates more than
the other strategies thus explaining the difference in performance. It is also
clear that CS only cooperates with itself and Handshake: it is a very aggressive
strategy.

>>>>

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a variety of strategic interactions. Each player chooses
simultaneously and independently
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as zero determinant (ZD) strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings (more details given in
Section~\ref{sec:the_moran_process}).
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, stewart2013extortion}, however these often make simplifying assumptions
or are limited to classes of strategies
such as memory-one strategies that only use the previous round of play.

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 50 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}).

@@

Using the Axelrod library and the many strategies it contains, we obtain
fixation probabilities for all pairs of strategies, identifying
those that are effective invaders and those resistant to invasion, for
population sizes $N=2$ to $N=14$. Moreover we present a number of strategies
that were created via reinforcement algorithms (evolutionary and particle
swarm algorithms) that are among the best invaders and resistors of invasion
known to date, and show that handshaking mechanisms naturally arise from these
processes as an invasion-resistance mechanism.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item The strongest resistors specifically evolve or have a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

@@

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider relatively simple strategies. Some work has
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

@@

To carry out this numerical experiment, \input{./tex/num_strategies}
strategies, listed (with their properties) in Appendix~\ref{app:list_of_players},
are used from the Axelrod library. There are
\input{./tex/num_stochastic.tex}stochastic and
\input{./tex/num_deterministic.tex}deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective, it differs from TF1 and TF2
in that it lacks a handshake mechanism. Figure~\ref{fig:tf3} shows
all 8 states of the strategy produced by the training process (states 3 and 8
are not reachable).

TF2 always starts with CD and will defect against opponents that start with
DD\@. It plays CDD against itself and then cooperates thereafter; Fortress3 and
Fortress4 also use a similar handshake and cooperate with TF2. Cooperation
can be rescued after a failed handshake by a complex sequence of plays
which sometimes results in mutual cooperation with Firm but Fair, Grofman, and
GTFT, and a few others with low probability. TF2 defects against all other
players in the study, barring
unusual cases arising from particular randomizations. Figure~\ref{fig:tf2} shows
all 16 states of the strategy (states 6 and 7 are not reachable).

@@

% Todo: verify 1000 repetitions

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item Defector: it always defects. Since it has no interactions with other
        defectors (recall that \(N=2\)), its aggressiveness is rewarded.
    \item Aggravater, which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake, a slightly less aggressive version of the Collective
        Strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself just as the Collective Strategy does. The
        difference is that CS will defect after the handshake if the opponent
        defects while handshake will not.
\end{enumerate}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/average_rank_vs_population_size_invade.pdf}
    \caption{\textbf{Invasion}: Ranks of all strategies according to \(x_1\) for different
    population sizes.}
    \label{fig:ranks_v_size_invade}
\end{figure}

@@

\begin{figure}[!hbtp]
    \centering
    \includegraphics[height=.9\textheight]{./img/average_rank_vs_population_size_resist.pdf}
    \caption{\textbf{Resistance}: Ranks of all strategies according to \(x_{N-1}\) for different
    population sizes.}
    \label{fig:ranks_v_size_resist}
\end{figure}

@@

As described in Section~\ref{sec:strong_invaders} the strategies trained with
the payoff maximizing objective are among the best invaders in the library
however they are not as resistant to invasion as the strategies trained using a
Moran objective function. These strategies include trained finite state machine
strategies, but they do not appear to have handshaking mechanisms. Therefore it
is reasonable to conclude that the objective function is the cause of the
emergence of handshaking mechanisms. More specifically, TF1 and TF2 evolved
handshake for high invasion resistance. TF3 is a better total payoff maximizer
which makes it a better invader along with the strategies
trained to maximize total payoff since successful fitness proportionate selection
is necessary for invasion. Training with an objective with initial population
mix other than $(N/2, N/2)$ may favor invasion or resistance.

@@

For all starting populations
\(i\in\{1, N/2, N-1\}\) the ranks of strategies are relatively stable across the
different values of \(N>2\) however for \(N=2\) there is a distinct difference.
This highlights that there is little that can be inferred about the evolutionary
performance of a strategy in a large population from its performance in a small
population. This is confirmed by the performance of the zero determinant strategies: while
some do rank relatively highly for \(N=2\) (ZD-extort-4 has rank 16) this rank
does not translate to larger populations.

@@

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than always cooperate necessarily involves
a defection. (A handshake consisting of a long sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1\_1\_1 to become better resistors
they need a longer or more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders need to maximize match payoff to benefit
from fitness proportionate selection,
and so in the absence of a handshake mechanism, knowledge of the population
distribution, or some identifying label on the opponent,
a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping dramatically
as the population size increases.

@@

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only revealed for CS (a human designed strategy) but also for
two FSM strategies (TF1 and TF2) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. Figure~\ref{fig:cooperation_rates} shows the cooperation
rate of TF1, TF2, TF3 and CS
for each round of a match against all the opponents in this study. While TF3
does not have a strict handshake mechanism it is clear that all these strategies
start a match by cooperating. It is then evident that TF3 cooperates more than
the other strategies thus explaining the difference in performance. It is also
clear that CS only cooperates with itself and Handshake: it is a very aggressive
strategy.
